Quantifying stochastic noise in cultured\\ circadian reporter cells

### 2015-05-06 response_letter.tex

Figure S7 indicated that although it was possible to differentiate between noise sources when only one type of noise was present, determining changes to one noise source when both are present is likely not identifiable from the data.
Ultimately, however, we don't believe that the key insight of our method is determining the explicit source of the stochastic variation.
Since both types of noise contribute to determining the population-level dynamics, quantifying the overall stochastic noise is the main goal.
However, based on the evidence presented in this study and in previous studies, we offer the explanation that cell-autonomous noise is likely the principal source of rhythm damping.

@@

\begin{reviewer}
M2. The method is of potential use for many labs including my own. What I was missing, however, are practical guidelines: How many samples are needed, which sampling rates, etc. are required for the method to work and yield ? How does it scale with increasing or decreasing sampling rates? Importantly, how do I decide whether a possible change in damping and thus noise level is significant, can one bootstrap this? Should one always use the fitting procedure with matrix pencil method, Hilbert transform to align phases, etc.? Here, a few trials on simulated time-series may suffice to provide such information.
\end{reviewer}

This is an excellent suggestion, and we have added additional sections to the manuscript describing practical considerations for experimental results.
More important than the sampling rate is the overall duration of the experiment, as this allows a more confident regression of the amplitude window.

The following paragraph was added to the discussion to address practical considerations in experimental design and method choice:

\begin{manuscript}{Page 15}
 In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
 {\bfseries
An overview of the computational steps involved in this method are outlined in Figure~S8.
While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
We also note that there are many available tools for detrending and regressing time-series data.
While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.
}
\end{manuscript}

Additionally, we have added a supplemental figure, Figure S8, to provide a more detailed overview of the method:

\begin{manuscript}{Page 23}
  \begin{center}
  \includegraphics[width=0.8\textwidth]{figures/pdfs/flowchart_schematic.pdf}
  \end{center}
{\bfseries Figure S8: Overview of the computational method.} ({\bfseries A}) Raw bioluminescence traces are collected from cultured cells. Since the method is mainly dependent on amplitude over time, longer experiments yield more identifiable damping rates. ({\bfseries B}) Raw traces are detrended and denoised. This can be accomplished using a variety of methods, including the discrete wavelet transform. ({\bfseries C}) The window of the oscillations (amplitude vs.\ time) is plotted on a semi-log axis to make the exponential damping apparent. Window functions can be calculated using a Hilbert transform, or simply by using the absolute value of the local peaks (as was done here). A linear regression of the window function in semi-log space will yield both the initial amplitude (intercept) and damping rate (slope), with care being taken to avoid the over-influence of outlier points. Standard statistical techniques could then be used to verify that differences in slope are indeed significant.
\end{manuscript}

We elected not to include simulated time-series demonstrations of these points, mainly because of the difficulty in getting simulated time-series to match the variability seen in biological systems.
Because of this, we believed recommended sampling rates and durations would not necessarily be reflective of biological data.

In regards to using the Hilbert transform to align phases, this was done mainly to be able to analyze the dephasing kinetics from a single-cell perspective.
Since there are relatively few sources of single-cell circadian data, this method is likely not applicable to many external studies.

%

% I'll add a paragraph on practical considerations for implementing the algorithm. Sampling rate is less important than overall recording duration. There are many ways to do the nonlinear regression and analysis, which likely depends on the data being analyzed. For looking specifically at the damping rate, the best method is likely detrending, followed by a semilog plot of window vs.\ time, from which more traditional statistical techniques on comparing two regressions (ancova, bayesian inference)

@@

\begin{reviewer}
Minor concerns:

Relating to question M1, the authors refer to their previous work, ref. 23, where it was shown that phase drift leads to an exponentially decreasing population signal. Otoh, period heterogeneity would lead to a time-squared exponent; where is the latter formula derived? If this was done by Rougemont and Naef, they should be cited.
\end{reviewer}

We have largely re-worded this section of the text in response to the comments of other reviewers. Despite this, we have added a further citation to Rougement \& Naef, as they present a more detailed theoretical analysis of the windowing function:

\begin{manuscript}{Page 21}
   ({\itshape Right}) A deterministic, heterogeneous population dephases due to different free-running periods. In this case, the population displays ballistic phase diffusion, in which envelope of the population-averaged expression changes proportionally to $t^2$. {\bfseries (See Rougemont \& Naef, 2007 (22) for a more detailed discussion).}
\end{manuscript}

In general, however, the concept that Brownian particles diffuse with a variance that scales with $t$, and that ballistic particles diffuse with a variance that scales with $t^2$, is very old. The result presented here follows from the connection between amplitude and the variance of the phase probability.

@@

\begin{reviewer}
On p. 3, comments regarding reference 17: these authors rather submitted that either damped or self-sustained oscillators describe the data equally well, which actually more justifies the present method, which assumes self-sustained oscillations.
\end{reviewer}

We thank the reviewer for catching this inconsistency, have have updated the text accordingly:

\begin{manuscript}{Page 3}
  Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, {\bfseries capture the noise characteristics seen in single-cell fibroblast data equally well as limit-cycle oscillators (20).}
\end{manuscript}

However, we note that our assumptions and the concept of noise-driven oscillators are not necessarily exclusive. In the cited work, the authors demonstrate that given a constant noise input, the damped oscillator filters the noise to amplify noise in the circadian frequencies. This yields an oscillator with similarly sustained amplitudes to a noisy limit-cycle.

@@

\begin{reviewer}
The method concerns primarily detection of noise levels, in the introduction it is emphasized that an amplitude boosting may ultimately be desirable in e.g. a clinical setting. Introduction would be strengthened by also including more use cases for the detection of noise strength.
\end{reviewer}

In addition to expanding the discussion on cellular noise in response to a previous reviewer's comments, we have added some additional remarks on the importance of measuring stochastic noise:

\begin{manuscript}{Page 3}
  {\bfseries
  The amount of noise in system is therefore linked to the ability of tissue-level clocks to maintain high-amplitude rhythms.

Despite the averaging that occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator. Noise in circadian rhythms has long been considered an important factor in how circadian rhythms have evolved (18).} A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues (19).
\end{manuscript}

We also note that the importance of amplitude boosting is linked to the measurement of noise, as noise likely plays a key role in determining the rate at which rhythms damp in peripheral tissues.

@@

\begin{reviewer}
For the discussion: authors argue that the amplitude/noise ratio seem to have evolved to something close to an optimum. Yet, in the paper by Liu et al. in Cell 129, 2007, Cry2 knockouts seemed to have even stronger rhythms?
\end{reviewer}

This is an interesting point. We are careful in saying that there may indeed be perturbations or knockouts which shift the clock to a region of higher stability, but that on average perturbations tend to lead to less robust rhythms.

We sadly don't have access to the raw data of Liu et al.\ to look more closely at their observations. In the genome-wide siRNA screen, however, Cry2 knockdown was one of the several positive controls run in each plate.
From this data, it seems that Cry2 knockdown seems to simultaneously reduce both the population-level amplitude and damping rate:

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=0.7\textwidth]{response_figs/cry2_ts.pdf}
  \end{center}
  \caption{{\bfseries Bioluminescence trajectories of control and Cry2 knockdown wells.} ({\itshape Left}) Control wells which did not contain an siRNA perturbation. ({\itshape Right}) Wells with Cry2 siRNA.}
\end{figure}

Looking at the parameter fits of these data, we see that while Cry2 siRNA does indeed seem to make the oscillators less noisy (lower damping rate), it also slightly reduces their amplitude. These results are mirrored by the dose-dependent Cry2 knockdown from the same study (Zhang et al., Cell 2009), which shoes a reduction of amplitude at higher Cry2 si concentrations.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=0.4\textwidth]{response_figs/cry2_dist.pdf}
  \end{center}
  \caption{{\bfseries Distribution of amplitude and damping rate fits for control and Cry2 siRNA wells.} The mean of both distributions is highlighted with a black dot.}
\end{figure}

However, since these points require comparing population-level amplitudes to single-cell level amplitudes (which may not be directly comparable), we decided to leave it out of the manuscript.

@@

\section*{Reviewer \#4}

\begin{reviewer}
Circadian rhythms play important roles in maintaining biological homeostasis. Recent research suggests that environmental conditions, like night work or an unusual feeding schedule, may have important implications for developing metabolic disease. Therefore it is of interest to identify therapeutic means to manipulate the circadian rhythms. Here, St. John and Doyle explore the relationship between stochastic noise and circadian oscillations, as typically observed at the cell population level and suggest a method to infer stochastic noise properties from population behavior. Overall, the work seems too tightly focused on a narrow scientific niche and the method for deducing noise properties is not convincing. The major criticisms are as follows:

1. The relationship between stochastic noise or cell-to-cell variability and how that impacts a population response is not particularly new. The observation that cell to cell variability gives rise to a dampened response at the cell population level has been observed before for other signaling systems, such as NF-kappaB (e.g., Nelson et al. Science (2004) 306:704-708). It is then unclear as to what really is new here.
\end{reviewer}

We agree that the relationships between cell synchrony, stochastic noise, and damped oscillations have been previously explored in many fields, including circadian rhythms. In addition to a more through description of noise sources, we have included an explicit reference to the presence of these phenomena in other systems in the introduction:

\begin{manuscript}{Page 3}
  {\bfseries This type of behavior has also been seen in
  other experimental systems, such as NF-$\kappa$B signaling or yeast glycolytic oscillations (16, 17).}
\end{manuscript}

The novel contribution of our manuscript is that we can use changes in the damping rates of population-level data to gain new insight into the relative levels of noise present under different perturbations.
From the datasets we analyze, damping appears to be independently altered (from period, amplitude) by genetic perturbations, and therefore likely provides new insight into the underlying oscillator.
Our main message is that the damping rate is due to cellular noise, but that the damping rate contains information about the oscillator which is currently going unused.

@@

\begin{reviewer}
2. The authors use a leap of logic to justify their approach in extracting stochastic noise parameters from cell population data. First the authors show that by fitting a collection of models to single cell data and then averaging these mathematical models together, they can predict that the cell population response will become dampened when there is significant variability in the single cell responses (inductive reasoning?). Based on this demonstration, they then develop an approach that does the opposite: from population data, they try to deduce the values for the stochastic noise properties (deductive reasoning?). The problem seems to be that there could be other explanations for why miRNAs within a library affect a signaling network, such as miRNAs limit the efficiency of the biochemical signaling network that underpins the oscillatory behavior. Overall, the approach is not convincing in providing the claimed insight. 
\end{reviewer}

As in all modeling work, we must make simplifying assumptions that may be drawn from incomplete experimental data.
In developing our approach for inferring noise properties from population-level data, we have invoked two main assumptions: that individual cells are both independent and sustained oscillators.
We are careful to list these assumptions in the introduction, as we agree that the validity of our results rests upon whether or not these assumptions hold in the systems under consideration.
However, to the best of our knowledge, there is relatively strong experimental support to both of these assumptions in cultured circadian reporter cells.
In the earlier parts of our paper, we are demonstrating that single-cell noise is {\itshape sufficient} to explain the observed damping behavior, and indeed quantitatively accurate predictions for population-level damping rate are obtained from a simple model not fit to the noise properties.

In regards to other potential explanations for the effects of siRNA, we suspect intercellular signalling networks do not play a significant role in cultured cell oscillations (as we assume each cell is independent).
These perturbations are thought to affect the cell-autonomous transcription-translation feedback loop, resulting in potentially altered noise properties.
Indeed, the siRNA method has been well-accepted by the circadian community, and are widely used to study the effect of gene knockdowns on amplitude and period.

@@

\begin{reviewer}
Minor comments:
1. ``which'' is used incorrectly pretty much throughout the manuscript. should be ``that''
\end{reviewer}

We thank the reviewer for pointing out this error. We have corrected our usage throughout the manuscript.
